---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r installlibraries, echo = TRUE}

        install.packages('dplyr', repos="http://cran.rstudio.com/")
        library(dplyr)
        install.packages('dtplyr', repos="http://cran.rstudio.com/")
        library(dtplyr)
        install.packages('ggplot2', repos="http://cran.rstudio.com/")
        library(ggplot2)




        workingdir <- getwd()
        
        destfile <- paste(workingdir, "/activity.csv", sep = "")
        if (!file.exists(destfile)) {
                print("got here")
                #  messge stub for a possible error handler
                print(paste("activity.csv ", 
                            "does not exist in the working direct ",
                            "- please put it there"))
                
        } else {

                activity <- read.csv("activity.csv" , header = TRUE)
                tblactivity <- tbl_df(activity)
                rm(activity)
        }

```




## What is mean total number of steps taken per day?
### Make a histogram of the total number of steps taken each day
``` {r, echo=TRUE}
        tblsteps <- tblactivity %>%
        select(steps,
               date,
               interval) %>%
        filter(!is.na(steps))%>%
        mutate(intervaldate = as.Date(date, format = "%Y-%m-%d")) %>%
        arrange(intervaldate) %>%
        group_by(intervaldate) %>%
        summarise(daysteps = sum(steps))

        plot1 <- ggplot(tblsteps, aes(daysteps))
        plot1 <- plot1 + geom_histogram(bins = 10, aes(fill = ..count..))
        plot1 <- plot1 + labs(x = "Number of Steps per Day" ,
                              y = "How Many Days at Each Count", 
                              title = "Distribution of Number of Steps per Day",
                              subtitle = "2012")
        plot1 <- plot1 + theme(plot.title=element_text(size=14, 
                                                       hjust=0.5, 
                                                       face="bold", 
                                                       colour="darkorchid4", 
                                                       vjust=-1)) 
 
        
        print(plot1)
```

### Calculate and report the mean and median total number of steps taken per day
```{r, echo = TRUE}
        tblstats <- tblactivity %>%
                select(steps,
                       date) %>%
                filter(!is.na(steps)) %>%
                group_by(date) %>%
                summarise(totalsteps = sum(steps))
        
        tblstats <- tblstats %>%
                select(totalsteps) %>%
                summarise(meansteps = mean(totalsteps), mediansteps = median(totalsteps))
        
        strmean <- paste("The mean total number of steps take per day = ",
                         tblstats$meansteps)
        strmedian <- paste("The median total number of steps take per day = ",
                         tblstats$mediansteps)
        print(strmean)
        print(strmedian)

```        

## What is the average daily activity pattern?
### Make a time series plot  of the 5-minute interval x-axis,
### and the average number of steps taken, averaged across all days y-axis

```{r, echo = TRUE}
        tblintervals <- tblactivity %>%
                select(steps,
                       interval,
                       date) %>%
                filter(!is.na(steps)) %>%
                arrange(interval) %>%
                group_by(interval) %>%
                summarise(intervalmeansteps = mean(steps))

        # get the interval with the highest mean of steps
        tblmaxinterval <- tblintervals %>%
                        arrange(intervalmeansteps)
        
        # find the max mean step interval
        maxrow <- nrow(tblmaxinterval)
        maxinterval <- tblmaxinterval[maxrow, 1]
        maxmeansteps <- tblmaxinterval[maxrow,2]
        
        # get the interval dates
        tbldates <- tblactivity %>%
                select(date) %>%
                mutate(intervaldate = as.Date(date, format = "%Y-%m-%d")) %>%
                arrange(intervaldate) %>%
                summarise(mindate = min(intervaldate), maxdate = max(intervaldate))

        
        mindate <- tbldates[1,1]
        maxdate <- tbldates[1,2]
        
        # plot the histogram of steps each day
        plot3 <- ggplot(tblintervals, aes(x =interval, 
                                      y = intervalmeansteps))
        plot3 <- plot3 + geom_line(size=1)
        plot3 <- plot3 + labs(x = "Interval Number" ,
                              y = "Average Steps per Interval", 
                              title = "Average Steps for Each Interval Across All Days",
                              subtitle = paste("Dates from ", 
                                                format(mindate, format="%Y-%m-%d"),
                                                " to ",
                                                format(maxdate, format="%Y-%m-%d")))
        plot3 <- plot3 + theme(plot.title=element_text(size=14, 
                                                       hjust=0.5, 
                                                       face="bold", 
                                                       colour="darkorchid4", 
                                                       vjust=-1))
        plot3 <- plot3 + theme(plot.subtitle=element_text(size=10, 
                                                          hjust=0.5, 
                                                          face="bold", 
                                                          colour="black", 
                                                          vjust=-1))
        plot3 <- plot3 + theme(panel.border = element_rect(colour = "black",
                                                           fill=NA, 
                                                           size=2))
        submsg <- " <- Highest average interval "
        submsg <- paste(submsg, maxinterval)
        submsg <- paste(submsg, " with a mean of ")
        submsg <- paste(submsg, format(maxmeansteps, nsmall=3))
        plot3 <- plot3 + annotate("text", x = 1550, y = 205, label = submsg, color = "blue4")
        
        
        print(plot3)
```


## Imputing missing values
### Calculate and report the total number of missing values in the dataset
### (i.e. the total number of rows with NAs)
```{r, echo = TRUE }        
        tblNAs <- tblactivity %>%
        select(steps,
               interval,
               date) %>%
        filter(is.na(steps))

        numNAs <- nrow(tblNAs)
        strmsg <- "The total number of missing values in the dataset is "
        print(paste(strmsg, numNAs))
````



## Are there differences in activity patterns between weekdays and weekends?
