---
title: "Reproducible Data - Assignment 1"
author: "Vicki Cocco"
date: "March 28, 2017"
output: html_document
---

```

## Project Description
=======================

### Install r Libraries
```{r installlibraries, echo = TRUE}
        install.packages('dplyr', repos="http://cran.rstudio.com/")
        library(dplyr)
        install.packages('dtplyr', repos="http://cran.rstudio.com/")
        library(dtplyr)
        install.packages('ggplot2', repos="http://cran.rstudio.com/")
        library(ggplot2)
        install.packages('lubridate', repos="http://cran.rstudio.com/")
        library(lubridate)
        install.packages('knitr', repos="http://cran.rstudio.com/")
        library(knitr)
```
### Read the activity file
```{r readingactivity, echo = TRUE}
        
        workingdir <- getwd()
        
        destfile <- paste(workingdir, "/activity.csv", sep = "")
        if (!file.exists(destfile)) {
                print("got here")
                #  messge stub for a possible error handler
                print(paste("activity.csv ", 
                            "does not exist in the working direct ",
                            "- please put it there"))
                
        } else {

                activity <- read.csv("activity.csv" , header = TRUE)
                tblactivity <- tbl_df(activity)
                rm(activity)
                }
```
            